Conceptualizing open health data platforms for low- and middle income countries

### 2024-03-25 index.tex
Remove project, revert to single page html

A commonly used perspective to scrutinize digital health is to consider
it as a digital platform \citep{dereuver2018digital}. Digital platforms
have disrupted many sectors but have just started to make inroads into
highly regulated industries such as healthcare \citep{ozalp2022digital}.
In this light, the challenges faced by LMICs in establishing national
digital health platforms have a lot in common with those faced by high
income countries. From a technological perspective, interoperability
issues, weak integrations, siloed data repositories and overall lack of
openness are often reported as key impediments
\citep{malm-nicolais2023exploring, mehl2023fullstac}. From a societal
perspective, issues pertaining to the winner-takes-all nature of digital
platforms are hotly debated as many jurisdictions make work to ensure
these platforms indeed serve the common good of achieving universal
health coverage \citep{sharon2018when}.

@@

Particularly relevant in the context of health data platforms (HDPs), is
the conceptualization of openness. The objective of openness, the ways
to realize openness and the risks of opening up can be considered as
elements within the ongoing debate on data spaces
\citep{otto2022designing} and data solidarity
\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond} in
the healthcare domain, which is a current issue for high income
countries and LMICs alike. Openness is particularly relevant if we are
to realize a solidarity-based approach to health data sharing that i)
gives people a greater control over their data as active decision
makers; ii) ensures that the value of data is harnessed for public good;
and iii) moves society towards equity and justice by counteracting
dynamics of data extraction \citep{prainsack2022data}.

@@

This paper is motivated by the conflation of a number of developments
relevant to the design and implementation of solidarity-based HDPs in
LMICs. First, the OpenHIE framework \citep{openhie} has been adopted by
many sub-Saharan African countries \citep{mamuye2022health} as the
architectural blueprint for implementing nation-wide health information
exchanges (HIE), including Nigeria \citep{ohienigeria}, Kenya
\citep{khisif} and Tanzania \citep{tzhea2020}. These countries have, as
a matter of course, extended the framework to include ``analytics
services'' as an additional domain. The rationale for this addition is
to facilitate primary and secondary re-use of health data for academic
research, real-world evidence studies etc. which can be framed within
the context of ongoing efforts towards Findable, Accessible,
Interoperable and Reusable (FAIR) sharing of health data
\citep{guillot2023fair}. In doing so, however, we have implicitly moved
from conceptualizing digital health platforms (the original OpenHIE
specification) to health data platforms. This is problematic because the
notion of openness, which is assumed to be essential in establishing
solidarity-based approaches to data sharing, is inherently different for
a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods\citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for data sharing
and re-use. Distinguishing four types of data sharing
(Table~\ref{tbl-types-data-sharing}), the purpose of this paper is to
investigate how new standards and technologies that can establish
openness of health data platforms can be integrated into the OpenHIE
architecture framework. The lack of detailed specifications and
consensus of this addition to OpenHIE currently stands in the way of
development projects that aim to establish HDPs in LMICs. In addition,
we explicitly address the requirement of downward scalability as we aim
to implement health data platforms in resource constrained setting of
LMICs.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5600}}@{}}
\caption{Types of data sharing and in relation to new standards and
technology enablers to create
openness.}\label{tbl-types-data-sharing}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{New standards and technology enablers to create openness}
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{New standards and technology enablers to create openness}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{1} & Data at the most granular (patient) level, which is
persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
API} which has by now been incorporated in all major FHIR
implementations the FHIR standard can be readily used to support type 1
data sharing to patient-level data across a patient population
\citep{mandl2020push, jones2021landscape}. \\
\textbf{2} & Aggregated data, for example statistics for policy
evaluation and benchmarking & \textbf{SQL-on-FHIR specification}
\citep{sql-on-fhir}: provides a standardised approach to make FHIR work
well with familiar and efficient SQL engines that are most commonly used
in analytical workflows. Builds on FHIRPath \citep{fhirpath} expressions
in a logical structure to specify things like column names and unnested
items. Implementations of this approach are available or forthcoming,
including open source implementations such as Pathling
\citep{grimes2022pathling} and commercial offerings like Aidbox
\citep{aidbox}. \\
\textbf{3} & Data analytics modules, that provide access to work and
access the data. & \textbf{Federated learning (FL)}
\citep{rieke2020future} and \textbf{privacy- enhancing technologies
(PETs)} \citep{scheibner2021revolutionizing, jordan2022selecting}: new
paradigms that address the problem of data governance and privacy by
training algorithms collaboratively without exchanging the data itself.
Models can be trained on combined datasets and made available as open
source artifacts for decision support. Data analysts can use FL and PETs
to work with the data in a collaborative, decentralized fashion. \\
\textbf{4} & Trained models that have been derived from the data and can
be used stand-alone for decision support. & \textbf{ONNX} \citep{onnx}:
an open format built to represent machine learning models. \\
\end{longtable}

@@

In the paper, we present a design that extends the OpenHIE specification
to include the four types of data sharing mentioned above. Using the
full-STAC approach \citep{mehl2023fullstac} we combine open standards,
open technologies and open architectures into a coherent modular HDP
platform that can be configured and re-used across a variety of
use-cases. Subsequently, we employ a formative, naturalistic evaluation
to assess the technical risk and efficacy of the design
\citep{venable2016feds}. Given that it is prohibitively expensive to
evaluate with real users and real systems in the real setting, we aim to
minimize technological risks and maximize the efficacy of the design by
considering three real-world examples of health data platoform in LMICs,
namely i) the OpenHIM platform
(\url{https://jembi.gitbook.io/openhim-platform/}); ii) the ONA Canopy
platform \url{https://ona.io/home/services/canopy/}; and iii) the
MomCare platform (\url{https://health-data-commons.pharmaccess.org}. As
part of our design research, we have taken a narrative approach in
surveying existing scientific studies on health data platforms, focusing
on the seminal reports and subsequently searching forward citations. In
addition, we have searched the open source repositories (most notably
GitHub) and the online communities (OpenHIE community, FHIR community)
to search for relevant open standards, technologies and architectures.
This paper should not be considered as a proper systematic review.

@@

The recent convergence to FHIR as the de facto standard for information
exchange has fuelled the development of OpenHIE. FHIR is currently used
both for routine healthcare settings\citep{ayaz2021fast} and clinical
research settings \citep{duda2022hl7, vorisek2022fast} and is
increasingly being used in LMICs as well. The guidelines and standards
of the African Union explicitly state FHIR is to be used as the
messaging standard \citep{2023african}. The FHIR-native OpenSRP platform
\citep{mehl2020open} has been deployed in 14 countries targeting various
patient populations, amongst which a reference implementation of the WHO
antenatal and neonatal care guidelines for midwives in Lombok, Indonesia
\citep{summitinstitutefordevelopment2023bunda, kurniawan2019midwife}. In
India, FHIR is used as the underlying technology for the open Health
Claims Exchange protocol specification, which has been adopted by the
Indian government as the standard for e-claims handling \citep{hcx}.
This range of utilizations showcase the standards' widespread
applicability. The proceedings of the OpenHIE conference 2023 attest to
the fact that FHIR and open source technologies are embraced as critical
enablers in implementing health information exchanges in LMICs
\citep{ohie23}.

However, despite the increased use of FHIR as a common data model,
various studies have investigated its merits and performance vis-a-vis
other healthcare standards. Comparisons between OpenEHR, ISO 13606, OMOP
and FHIR have been made
\citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data}.
A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR
concluded that i) these three standards are functionally and technically
compatible, and therefore can be used side by side; and that ii) each of
these standards have their strengths and limitations that correlate with
their intended use as summarized in the Table~\ref{tbl-comparison}.

@@

\subsection{Open architecture: extending OpenHIE framework with a
lakehouse data
platform}\label{open-architecture-extending-openhie-framework-with-a-lakehouse-data-platform}

Data management and analytics platforms have undergone significant
changes since the first generation of data warehouses were introduced.
Recent studies have shown that the current practice has converged
towards the lakehouse as one of the most commonly used solution designs
\citep{armbrust2021lakehouse, hai2023data, harby2022data}. Lakehouses
typically have a zonal architecture \citep{hai2023data} where data is
ingested from the source systems in bulk (E), delivered to storage with
aligned schemas (L) and transformed into a format ready for analysis
(T). The discerning characteristic of the lakehouse architecture is its
foundation on low-cost and directly-accessible storage that also
provides traditional analytical DBMS management and performance features
such as ACID transactions, data versioning, auditing, indexing, caching,
and query optimization \citep{armbrust2021lakehouse}. Lakehouses thus
combine the key benefits of data lakes and data warehouses: low-cost
storage in an open format accessible by a variety of systems from the
former, and powerful management and optimization features from the
latter.

@@

While the lakehouse architecture separates the concerns of compute and
storage, the composable data stack takes the separation of concerns is
taken one step further. A composable data system
(Figure~\ref{fig-composable-data-stack}), not only separates the storage
(layer 3) and execution (layer 2), but also separates the user interface
(layer 1) from the execution engine by introducing standards for
Intermediate Representation (standard A) and Connectivity (standard B).
The composable data stack can be implemented with current open source
technologies (Figure~\ref{fig-cds-examples}). As an example, the Ibis
user interface is currently sufficiently mature to offer a standardized
dataframe interface to 19 different execution engines.

@@

\includegraphics{images/composable-data-stack-implementation.png}

@@

The premise of separating the user interface from the execution engine
is directly related to the key objective of the SQL-on-FHIR project
(\url{https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/}), namely to make
large-scale analysis of FHIR data accessible to a larger audience,
portable between systems and to make FHIR data work well with the best
available analytic tools, regardless of the technology stack. However,
as FHIR is represented as a graph of resources, with detailed semantics
defined for references between resources, data types, terminology,
extensions, and many other aspects of the specification; to use FHIR
effectively analysts require a thorough understanding of the
specification. Most analytic and machine learning use cases require the
preparation of FHIR data using transformations and tabular projections
from its original form. The task of authoring these transformations and
projections is not trivial and there is currently no standard mechanisms
to support reuse.

@@

Our proposal for extending HIE is as follows.

\begin{itemize}
\tightlist
\item
  add diagram, explicitly adding the component
\item
  fit in structure of OpenHIE specification
\item
  check which workflows are related to analytics
\end{itemize}

Following Hai we take a subset of the core functionalities of a data
lake. Because the SHR is the key source, don't need full stack of
ingestion of various sources.

@@

\begin{itemize}
\tightlist
\item
  Query-driven data discovery
\end{itemize}

@@

>>>>

A commonly used perspective to scrutinize digital health is to consider
it as a digital platform \citep{dereuver2018digital}. Digital platforms
have disrupted many sectors but have just started to make inroads into
highly regulated industries such as healthcare \citep{ozalp2022digital}.
In this light, the challenges faced by LMICs in establishing national
digital health platforms have a lot in common with those faced by high
income countries. From a technological perspective, interoperability
issues, weak integrations, siloed data repositories and overall lack of
openness are often reported as key impediments
\citep{malm-nicolais2023exploring, mehl2023fullstac}. From a societal
perspective, issues pertaining to the winner-takes-all nature of digital
platforms are hotly debated as many jurisdictions make work to ensure
these new digital health platforms indeed serve the common good of
achieving universal health coverage \citep{sharon2018when}.

@@

Particularly relevant in the context of health data platforms (HDPs), is
the conceptualization of openness. The shift in perspective from digital
platforms to data platforms coincides with the paradox of open
\citep{keller2021paradox}. Originally, openness of digital platforms
focused on open source, open standards and copyrights, which by has been
superseded by ``\ldots{} conflicts about privacy, economic value
extraction, the emergence of artificial intelligence, and the
destabilizing effects of dominant platforms on (democratic) societies.
Instead of access to information, the control of personal data has
emerged in the age of platforms as the critical contention.''
\citep{keller2021paradox}. These conflicts are particularly salient in
the healthcare domain, as is evidenced by polemic surrounding data
spaces \citep{otto2022designing} and data solidarity
\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond}.
Openness is particularly relevant if we are to realize a
solidarity-based approach to health data sharing that i) gives people a
greater control over their data as active decision makers; ii) ensures
that the value of data is harnessed for public good; and iii) moves
society towards equity and justice by counteracting dynamics of data
extraction \citep{prainsack2022data}.

@@

This paper is motivated by the conflation of a number of developments
relevant to the design and implementation of solidarity-based HDPs in
LMICs. First, the OpenHIE framework \citep{openhie} has been adopted by
many sub-Saharan African countries \citep{mamuye2022health} as the
architectural blueprint for implementing nation-wide health information
exchanges (HIE), including Nigeria \citep{dalhatu2023paper}, Kenya
\citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}. These
countries have, as a matter of course, extended the framework to include
``analytics services'' as an additional domain. The rationale for this
addition is to facilitate secondary reuse of health data for academic
research, real-world evidence studies etc. which can be framed within
the context of ongoing efforts towards Findable, Accessible,
Interoperable and Reusable (FAIR) sharing of health data
\citep{guillot2023fair}. In doing so, however, we have implicitly moved
from conceptualizing digital health platforms (the original OpenHIE
specification) to health data platforms. This is problematic because the
notion of openness, which is assumed to be essential in establishing
solidarity-based approaches to data sharing, is inherently different for
a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods\citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for data sharing
and reuse. Distinguishing four types of data sharing
(Table~\ref{tbl-types-data-sharing}), the purpose of this paper is to
investigate how new standards and technologies that can establish
openness of health data platforms can be integrated into the OpenHIE
architecture framework. The lack of detailed specifications and
consensus of this addition to OpenHIE currently stands in the way of
development projects that aim to establish HDPs in LMICs. In addition,
we explicitly address the requirement of downward scalability as we aim
to implement health data platforms in resource constrained setting of
LMICs.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5600}}@{}}
\caption{Types of data sharing and in relation to new standards and
technology enablers to create
openness.}\label{tbl-types-data-sharing}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{1} & Data at the most granular (patient) level, which is
persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
API}: has by now been incorporated in all major FHIR implementations the
FHIR standard can be readily used to support data sharing to
patient-level data across a patient population
\citep{mandl2020push, jones2021landscape}. \\
\textbf{2} & Aggregated data, for example statistics for policy
evaluation and benchmarking & \textbf{SQL-on-FHIR specification}
\citep{sql-on-fhir}: provides a standardised approach to make FHIR work
well with familiar and efficient SQL engines that are most commonly used
in analytical workflows. Builds on FHIRPath \citep{hl72020fhirpath}
expressions in a logical structure to specify things like column names
and unnested items. Implementations of this approach are available or
forthcoming, including open source implementations such as Pathling
\citep{grimes2022pathling} and commercial offerings like Aidbox
\citep{aidbox}. \\
\textbf{3} & Data analytics modules, that provide access to work and
access the data. & \textbf{Federated learning (FL)}
\citep{rieke2020future} and \textbf{privacy- enhancing technologies
(PETs)} \citep{scheibner2021revolutionizing, jordan2022selecting}: new
paradigms that address the problem of data governance and privacy by
training algorithms collaboratively without exchanging the data itself.
Models can be trained on combined datasets and made available as open
source artifacts for decision support. Data analysts can use FL and PETs
to work with the data in a collaborative, decentralized fashion. \\
\textbf{4} & Trained models that have been derived from the data and can
be used stand-alone for decision support. & \textbf{ONNX} \citep{onnx}:
an open format built to represent machine learning models. \\
\end{longtable}

@@

In the paper, we present a design that extends the OpenHIE specification
to include the four types of data sharing mentioned above. Using the
full-STAC approach \citep{mehl2023fullstac} we combine open standards,
open technologies and open architectures into a coherent modular HDP
platform that can be configured and reused across a variety of
use-cases. Subsequently, we employ a formative, naturalistic evaluation
to assess the technical risk and efficacy of the design
\citep{venable2016feds}. Given that it is prohibitively expensive to
evaluate with real users and real systems in the real setting, we aim to
minimize technological risks and maximize the efficacy of the design by
considering three real-world examples of health data platoform in LMICs,
namely i) the OpenHIM platform
(\url{https://jembi.gitbook.io/openhim-platform/}); ii) the ONA Canopy
platform \url{https://ona.io/home/services/canopy/}; and iii) the work
conducted at PharmAccess Foundation as part of the MomCare programme
(\url{https://health-data-commons.pharmaccess.org}). As part of our
design research, we have taken a narrative approach in surveying
existing scientific studies on health data platforms, focusing on the
seminal reports and subsequently searching forward citations. In
addition, we have searched the open source repositories (most notably
GitHub) and the online communities (OpenHIE community, FHIR community)
to search for relevant open standards, technologies and architectures.
This paper should not be considered as a proper systematic review.

@@

The recent convergence to FHIR as the de facto standard for information
exchange has fuelled the development of OpenHIE. FHIR is currently used
both for routine healthcare settings \citep{ayaz2021fast} and clinical
research settings \citep{duda2022hl7, vorisek2022fast} and is
increasingly being used in LMICs as well. The guidelines and standards
of the African Union explicitly state FHIR is to be used as the
messaging standard \citep{2023african}. The FHIR-native OpenSRP platform
\citep{mehl2020open} has been deployed in 14 countries targeting various
patient populations, amongst which a reference implementation of the WHO
antenatal and neonatal care guidelines for midwives in Lombok, Indonesia
\citep{summitinstitutefordevelopment2023bunda, kurniawan2019midwife}. In
India, FHIR is used as the underlying technology for the open Health
Claims Exchange protocol specification, which has been adopted by the
Indian government as the standard for e-claims handling \citep{hcx}.
This range of utilizations showcase the standards' widespread
applicability. The proceedings of the OpenHIE conference 2023 attest to
the fact that FHIR and open source technologies are embraced as critical
enablers in implementing health information exchanges in LMICs
\citep{ohie2023unconference}.

Various studies have investigated the merits of FHIR and its performance
vis-a-vis other healthcare standards. Comparisons between OpenEHR, ISO
13606, OMOP and FHIR have been made
\citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data}.
A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR
concluded that i) these three standards are functionally and technically
compatible, and therefore can be used side by side; and that ii) each of
these standards have their strengths and limitations that correlate with
their intended use as summarized in the Table~\ref{tbl-comparison}.

@@

\subsection{Open architecture: extending OpenHIE with a composable data
stack}\label{open-architecture-extending-openhie-with-a-composable-data-stack}

\subsubsection{Evolving the data lakehouse to a composable data
stack}\label{evolving-the-data-lakehouse-to-a-composable-data-stack}

Data management and analytics platforms have undergone significant
changes since the first generation of data warehouses were introduced.
Recent studies have shown that the current practice has converged
towards the lakehouse as one of the most commonly used solution designs
\citep{armbrust2021lakehouse, hai2023data, harby2022data}. Lakehouses
typically have a zonal architecture that follow the
Extract-Load-Transform pattern (ELT) \citep{hai2023data} where data is
ingested from the source systems in bulk (E), delivered to storage with
aligned schemas (L) and transformed into a format ready for analysis
(T). The discerning characteristic of the lakehouse architecture is its
foundation on low-cost and directly-accessible storage that also
provides traditional analytical DBMS management and performance features
such as ACID transactions, data versioning, auditing, indexing, caching,
and query optimization \citep{armbrust2021lakehouse}. Lakehouses thus
combine the key benefits of data lakes and data warehouses: low-cost
storage in an open format accessible by a variety of systems from the
former, and powerful management and optimization features from the
latter.

@@

While the lakehouse architecture separates the concerns of compute and
storage, the composable data stack takes the separation of concerns is
taken one step further. A composable data system
(Figure~\ref{fig-composable-data-stack}), not only separates the storage
(layer 3) and execution (layer 2), but also separates the user interface
(layer 1) from the execution engine by introducing standards including
Substrait for Intermediate Representation (standard A) and Apache Arrow
(standard B and C). The composable data stack can be implemented with
current open source technologies (Figure~\ref{fig-cds-examples}). As an
example, the Ibis user interface is currently sufficiently mature to
offer a standardized dataframe interface to 19 different execution
engines.

@@

\includegraphics{images/composable-data-stack-implementation-2.png}

@@

\subsubsection{SQL-on-FHIR v2 as an intermediate representation for FHIR
data in tabular
format}\label{sql-on-fhir-v2-as-an-intermediate-representation-for-fhir-data-in-tabular-format}

The premise of separating the user interface from the execution engine
is directly related to the key objective of the SQL-on-FHIR project
(\url{https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/}), namely to make
large-scale analysis of FHIR data accessible to a larger audience,
portable between systems and to make FHIR data work well with the best
available analytic tools, regardless of the technology stack. However,
as FHIR is represented as a graph of resources, with detailed semantics
defined for references between resources, data types, terminology,
extensions, and many other aspects of the specification. To use FHIR
effectively analysts require a thorough understanding of the
specification. Most analytic and machine learning use cases require the
preparation of FHIR data using transformations and tabular projections
from its original form. The task of authoring these transformations and
projections is not trivial and there is currently no standard mechanisms
to support reuse.

@@

\subsubsection{Extending OpenHIE with a FHIR-based composable data
stack}\label{extending-openhie-with-a-fhir-based-composable-data-stack}

We propose to extend the OpenHIE architecture with a ``Data and
Analytics Services'' domain with 5 services by synthesizing the
architecture of Hai et al. \citep{hai2023data} and the composable data
stack \citep{pedreira2023composable} (Figure~\ref{fig-ohie},
Table~\ref{tbl-data-and-analysis-services}).

@@

\caption{\label{fig-ohie}Proposed extension of the OpenHIE architecture
that includes ``Data and Analytics Services'' as an additional service
domain.}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1852}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8148}}@{}}
\caption{Definition of Data and Analysis
Services}\label{tbl-data-and-analysis-services}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Service
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functional requirements
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Service
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functional requirements
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Ingestion & \begin{minipage}[t]{\linewidth}\raggedright
\begin{itemize}
\tightlist
\item
  Metadata extraction
\item
  Metadata modeling
\end{itemize}
\end{minipage} \\
Storage & \begin{minipage}[t]{\linewidth}\raggedright
\begin{itemize}
\tightlist
\item
  File-based blob storage
\end{itemize}
\end{minipage} \\
Query \& Processing & \begin{minipage}[t]{\linewidth}\raggedright
\begin{itemize}
\tightlist
\item
  SQL-on-FHIR Runner
\item
  Execution engine on tabular data as defined in composable data stack
\end{itemize}
\end{minipage} \\
Maintenance & \begin{minipage}[t]{\linewidth}\raggedright
\begin{itemize}
\tightlist
\item
  SQL-on-FHIR View defintions
\item
  Maintenance-related functions as defined by Hai et al.
\end{itemize}
\end{minipage} \\
Exploration & \begin{minipage}[t]{\linewidth}\raggedright
\begin{itemize}
\tightlist
\item
  Interactive computing environment \citep{granger2021jupyter}
\item
  Capability to participate as a node in a federative learning network,
  including MPC
\end{itemize}
\end{minipage} \\
\end{longtable}

\paragraph{Ingestion}\label{ingestion}

Default workflow is extraction of data from SHR using Bulk FHIR API.
Data contains metadata (incl.~FHIR versions) and fully qualified
semantics, for example, coding systems. Despite this, metadata
extraction and metadata modeling is still required to meet the FAIR
requirements. Issues that need to be solved by these services:

\begin{itemize}
\tightlist
\item
  To prepare for future updates of FHIR versions
\item
  Implement late-binding principle of having increasingly more specific
  FHIR profiles as bulk FHIR data propagates through lakehouse
\end{itemize}

\paragraph{Storage}\label{storage}

\begin{itemize}
\tightlist
\item
  from ndjson to parquet
\item
  possibly used delta lake for time versioning
\item
  separation of storage from compute not only for benefits of lower TCO,
  but also be ready for federated learning and MPC in future
\end{itemize}

\paragraph{Query \& Processing}\label{query-processing}

\begin{itemize}
\tightlist
\item
  fit in structure of OpenHIE specification
\item
  check which workflows are related to analytics
\item
  Hai calls this `Maintenance'
\end{itemize}

\paragraph{Maintenance}\label{maintenance}

\begin{itemize}
\tightlist
\item
  SQL-on-FHIR Views provide new standard to support mADX aggregate
  reporting
\item
  Maintenance-related functions remain the same
\item
  NB: orchestration falls under data provenance
\end{itemize}

\paragraph{Exploration}\label{exploration}

@@

\subsection{Federated learning and multiparty
computation}\label{federated-learning-and-multiparty-computation}

\begin{itemize}
\tightlist
\item
  data stations??!
\end{itemize}
